---
title: "TMESplit quickstart"
author: "Kate Petrenko"
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        toc: true
vignette: >
    %\VignetteIndexEntry{TMESplit quickstart}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Overview

`TMESplit` provides formal statistical testing for whether cell-type
coordination structure differs between biological groups (e.g. tumour vs
normal, responder vs non-responder). It decomposes a patient × cell-type
composition matrix into shared and group-specific non-negative programs,
and returns a three-level hierarchy of p-values (global, per-group,
per-program).

The algorithm is implemented in Python (`TMESplitpy`) and called from R
via [basilisk](https://bioconductor.org/packages/basilisk), which
provisions a sandboxed conda environment automatically on first use.

# Status

This release (`0.99.0`) is a **pre-release scaffold**. The API surface
is in place but the algorithmic bodies land in Phase 3.2b. Loading the
package and inspecting the S4 class work today:

```{r}
library(TMESplit)

res <- new("TMESplitResult")
res
```

The toy fixture used for cross-language parity tests ships in
`inst/extdata`:

```{r}
f <- system.file("extdata", "toy_composition.rds", package = "TMESplit")
toy <- readRDS(f)
vapply(toy, dim, integer(2))
```

# Planned API

```r
library(TMESplit)
library(SummarizedExperiment)

# (A) SummarizedExperiment (cell-types x patients)
res <- tmesplit(se, group_col = "condition",
                beta_loss = 0.5, n_perm = 2000)

# (B) SingleCellExperiment: aggregate first
comp <- aggregateToComposition(sce,
                               patient_col = "patient_id",
                               celltype_col = "celltype_L2")
res <- tmesplit(comp, group_col = "condition")

# (C) Raw matrix + group vector
res <- tmesplit(X, group = group_vec, beta_loss = 0.5)

programs(res)           # per-program DataFrame + BH-FDR
pvalue(res, "global")
plotPrograms(res)       # ComplexHeatmap
```

# Session info

```{r}
sessionInfo()
```
