---
title: "TMESplit analysis walkthrough"
author: "Kate Petrenko"
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        toc: true
vignette: >
    %\VignetteIndexEntry{TMESplit analysis walkthrough}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
RUN_PYTHON <- requireNamespace("reticulate", quietly = TRUE) &&
    tryCatch({
        proc <- basilisk::basiliskStart(TMESplit:::.tmesplit_env)
        basilisk::basiliskStop(proc)
        TRUE
    }, error = function(e) FALSE)
```

# Introduction

This vignette walks through a complete TMESplit analysis on simulated
data that mimics a typical scRNA-seq experiment: two biological groups
(e.g., responder vs non-responder), shared multicellular programs, and
one group-specific program.

# Simulating realistic composition data

```{r simulate, eval = RUN_PYTHON}
set.seed(123)
n_patients <- 30
n_celltypes <- 20

ct_names <- paste0("CellType_", seq_len(n_celltypes))
base_props <- rexp(n_celltypes, rate = 1)
base_props <- base_props / sum(base_props)

simulate_group <- function(n, base, noise_sd = 0.3) {
    mat <- matrix(NA, nrow = n, ncol = length(base))
    for (i in seq_len(n)) {
        log_props <- log(base + 1e-6) + rnorm(length(base), sd = noise_sd)
        props <- exp(log_props)
        mat[i, ] <- props / sum(props)
    }
    colnames(mat) <- ct_names
    rownames(mat) <- paste0("P", seq_len(n))
    mat
}

group_A <- simulate_group(n_patients, base_props)

base_B <- base_props
base_B[1:3] <- base_B[1:3] * 2.5
base_B[4:5] <- base_B[4:5] * 0.3
base_B <- base_B / sum(base_B)
group_B <- simulate_group(n_patients, base_B)

mat <- rbind(group_A, group_B)
rownames(mat) <- paste0("P", seq_len(2 * n_patients))
group <- rep(c("GroupA", "GroupB"), each = n_patients)
```

# Running TMESplit

```{r run, eval = RUN_PYTHON}
library(TMESplit)

res <- tmesplit(mat, group = group,
                beta_loss = 0.5,
                n_perm = 500,
                n_runs = 5,
                seed = 42,
                verbose = FALSE)
res
```

# Interpreting results

## Three levels of significance

```{r levels, eval = RUN_PYTHON}
cat("Level 1 (global):", pvalue(res, "global"), "\n")
cat("Level 2 (per-group):\n")
print(pvalue(res, "group"))
```

The hierarchical testing follows closed-testing discipline: Level 2
results are only interpretable if Level 1 is significant, and Level 3
only if Level 2 is significant for that group.

## Program structure

```{r programs-table, eval = RUN_PYTHON}
programs(res)
```

## Shared programs

```{r w-shared, eval = RUN_PYTHON}
W_sh <- wShared(res)
cat("Shared programs:", ncol(W_sh), "\n")
cat("Cell types:", nrow(W_sh), "\n")
```

## Group-specific programs

```{r w-specific, eval = RUN_PYTHON}
W_sp <- wSpecific(res)
for (g in names(W_sp)) {
    cat(g, ": ", ncol(W_sp[[g]]), " specific program(s)\n", sep = "")
}
```

# Visualization gallery

## Program loadings

```{r viz-programs, eval = RUN_PYTHON, fig.width = 10, fig.height = 8}
plotPrograms(res)
```

## Patient activities

```{r viz-activities, eval = RUN_PYTHON, fig.width = 8, fig.height = 5}
plotActivities(res)
```

## Significance overview

```{r viz-significance, eval = RUN_PYTHON, fig.width = 7, fig.height = 4}
plotSignificance(res)
```

## Cell-type network

```{r viz-network, eval = RUN_PYTHON, fig.width = 7, fig.height = 7}
plotNetwork(res)
```

# Parameter guidance

| Parameter | Default | When to change |
|-----------|---------|---------------|
| `beta_loss` | 0.5 | Rarely. 0.5 (Itakura-Saito) suits compositional data. Use 2.0 (Frobenius) only if data is not proportional. |
| `n_perm` | 2000 | Lower (100-500) for exploration; 2000+ for publication. |
| `n_runs` | 10 | NMF restarts. 10 is robust; increase to 20 if results feel unstable. |
| `combiners` | "cauchy" | "bonferroni" for conservative; "cauchy" (ACAT) is default and well-calibrated. |
| `seed` | 42 | Set for reproducibility. |

# Working with SingleCellExperiment

```{r sce-example, eval = RUN_PYTHON}
library(SingleCellExperiment)
library(SummarizedExperiment)

n_cells <- 5000
patients <- sample(paste0("P", 1:20), n_cells, replace = TRUE)
celltypes <- sample(ct_names[1:10], n_cells, replace = TRUE,
                    prob = base_props[1:10] / sum(base_props[1:10]))
counts <- matrix(rpois(n_cells * 50, 5), nrow = 50)

sce <- SingleCellExperiment(
    assays = list(counts = counts),
    colData = DataFrame(
        patient = patients,
        celltype = celltypes,
        condition = ifelse(patients %in% paste0("P", 1:10), "Ctrl", "Treat")
    )
)

comp <- aggregateToComposition(sce,
                               patient_col = "patient",
                               celltype_col = "celltype")

cat("Composition SE:", nrow(comp), "cell types x", ncol(comp), "patients\n")
cat("Proportions sum to 1:",
    all(abs(colSums(assay(comp)) - 1) < 1e-10), "\n")
```

# Session info

```{r}
sessionInfo()
```
