---
title: "TMESplit quickstart"
author: "Kate Petrenko"
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        toc: true
vignette: >
    %\VignetteIndexEntry{TMESplit quickstart}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
RUN_PYTHON <- requireNamespace("reticulate", quietly = TRUE) &&
    tryCatch({
        proc <- basilisk::basiliskStart(TMESplit:::.tmesplit_env)
        basilisk::basiliskStop(proc)
        TRUE
    }, error = function(e) FALSE)
```

# Overview

TMESplit tests whether cell-type coordination structure differs between
biological groups. Given a patient x cell-type composition matrix with
group labels, it:

1. Runs per-group NMF to discover multicellular programs
2. Matches programs across groups to identify shared vs group-specific ones
3. Applies a three-level hierarchical permutation test

# Quick example

## Load the toy fixture

```{r load-data}
library(TMESplit)

f <- system.file("extdata", "toy_composition.rds", package = "TMESplit")
toy <- readRDS(f)
str(toy)
```

## Run TMESplit

```{r run-tmesplit, eval = RUN_PYTHON}
mat <- rbind(toy$A, toy$B)
group <- rep(c("A", "B"), c(nrow(toy$A), nrow(toy$B)))

res <- tmesplit(mat, group = group,
                n_perm = 100, n_runs = 3, seed = 42, verbose = FALSE)
res
```

## Inspect results

```{r inspect, eval = RUN_PYTHON}
pvalue(res, "global")

pvalue(res, "group")

programs(res)
```

## Visualize

```{r plot-programs, eval = RUN_PYTHON, fig.width = 8, fig.height = 6}
plotPrograms(res)
```

```{r plot-activities, eval = RUN_PYTHON, fig.width = 7, fig.height = 4}
plotActivities(res)
```

```{r plot-significance, eval = RUN_PYTHON, fig.width = 7, fig.height = 4}
plotSignificance(res)
```

```{r plot-network, eval = RUN_PYTHON, fig.width = 7, fig.height = 7}
plotNetwork(res)
```

## Using SummarizedExperiment input

```{r se-dispatch, eval = RUN_PYTHON}
library(SummarizedExperiment)

se <- SummarizedExperiment(
    assays = list(proportions = t(mat)),
    colData = DataFrame(condition = group)
)

res_se <- tmesplit(se, group_col = "condition",
                   n_perm = 100, n_runs = 3, seed = 42, verbose = FALSE)

pvalue(res_se, "global")
```

# Session info

```{r}
sessionInfo()
```
